An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-06-17 main.tex
Walkthrough

Additionally, \Catra{} is able to efficiently generate a quantifier-free
Presburger formula form of such an instance of constraints and automata,
something beyond the reach of general model checkers like~\Nuxmv.

In general we contribute:
\begin{itemize}
\item A Calculus
\item QE-implementation
\end{itemize}

@@

>>>>

Additionally, \Catra{} is able to efficiently generate a quantifier-free
Presburger formula representing such an instance of constraints and automata,
something beyond the reach of general model checkers like~\Nuxmv.

In summary, we contribute:
\begin{itemize}
\item The \Calculus{} calculus to efficiently compute the Parikh image of products of Parikh automata.
\item A specialised method for quantifier elimination that blends well with \Calculus{} to efficiently compute quantifier-free Presburger representations of a given Parikh image.
\item Experiments illustrating the performance of \Calculus{} on real-world examples from string solving, including \NrBenchmarks{} instances in a standardised format made available for future studies.
\item The \Catra{} tool for solving such instances, containing an implementation of \Calculus{}, the over-approximation described in~\cite{approximate-parikh}, and an adapter for the~\Nuxmv{} model checker~\cite{nuxmv}.
\item Suggestions for how to efficiently implement \Calculus{} in a modern automated theorem prover, including strategies for case splitting, clause learning, and constraint propagation for connectedness.
\item Finally, a published description of the well-known bug in~\cite{generate-parikh-image} previously only described in folklore.
\end{itemize}

@@

Consider the two automata $\SomethingCSomething{}$
(Figure~\ref{fig:something-c-something}) and $\AcaOrBc{}$
(Figure~\ref{fig:aca-or-bc}). Assume we want to find a value in the Parikh image
of their product where $\TransitionCount{a} > \TransitionCount{b}$, that is find
a count of characters in a string accepted by both automata where there are more
a:s than b:s.

The na\"ive approach would be to first compute the product of the automata and
then compute the Parikh image of its product. However, we can take advantage of
the constraint to prune the automata before computing the product and help
reduce the blow-up. To give an intuition for the core idea of \Calculus{}, we
will annotate the transitions of $\SomethingCSomething$ and $\AcaOrBc$ with
variables illustrating how many times they are taken, and perform informal
reasoning on them corresponding to the reasoning performed by our calculus.

\begin{figure}[ht]
  \begin{minipage}[b]{0.75\linewidth}
  \centering 
    \includegraphics[width=\textwidth]{aca_or_bc}
    \caption{An automaton recognising the regular expression
    $\mathtt{ac^*a|b(bb)^*c}$.}\label{fig:aca-or-bc}

    \includegraphics[width=\textwidth]{something_c_something}
    \caption{An automaton recognising the regular expression
    $\mathtt{.^*c.^*}$.}\label{fig:something-c-something}
  \end{minipage}
  \end{figure}

Note that we have already performed some reasoning in the starting figures -- we
use the same symbol for the number of times each of the upper transitions of
$\AcaOrBc$ are taken, since we know that they must be taken the same number of
times, and we introduced the constant $1$ for the middle transition in
$\SomethingCSomething$, since it must always be taken precisely once to reach
the accepting state. We set the initial flow into the starting state to 1 and
use linear equations to enforce balanced flow throughout the automaton to
determine the relations between the counting variables on each transition (e.g.
$1 + {l_b}' = l_a + l_b$, or $l_a + l_c' = 1$). Using standard linear equation
reasoning loses additional variables and gives us the annotations in
Figure~\ref{fig:example-simplify-1}.

\begin{figure}[h]
  \begin{minipage}[b]{0.75\linewidth}
  \centering 
    \includegraphics[width=\textwidth]{aca_or_bc_simplified}
    \caption{$\AcaOrBc{}$ with its symbolic transition counters after
    simplification of their equations.}\label{fig:example-simplify-1}
  \end{minipage}
  \end{figure}

  We now need to relate the number of times a transition was taken to the
  character counts we are really interested in. Note that this also puts the two
  automata into contact with each other, before any product has been computed.
  Some consideration leads us to the definition that the number of occurrence of
  a each character corresponds to the sum of usages of transitions containing
  it, e.g. relations of the kind $\TransitionCount{a} = 2 - 2l_c$, and
  $\TransitionCount{b} = 2 l_b' + l_c'$.\footnote{For the attentive reader: in this
  example, we leave the catch-all transitions of $\SomethingCSomething$
  unconstrained since they will be taken care of by the product and would
  complicate the reasoning since we cannot yet know which character they will
  correspond to in the final product. In other words, this is a safe
  under-approximation that only limits our ability to reason before
  computing the product.}

This relation can be combined with the constraint that $\TransitionCount{a} >
\TransitionCount{b}$ to obtain $2 - 2l_c > 2 l_b' + l_c'$. Combining this
equation with the requirement from $\SomethingCSomething$ that
$\TransitionCount{c} \geq 1$ and $\TransitionCount{c} = l_c + l_c'$ from
$\AcaOrBc$ gives $0 \geq l_b'$, and since we cannot use a transition a negative
number of times $l_b' = 0$, which gives the automaton in
Figure~\ref{fig:example-simplify-2}.

\begin{figure}[h]
  \begin{minipage}[b]{0.75\linewidth}
  \centering 
    \includegraphics[width=\textwidth]{aca_or_bc_simplified_2}
    \caption{$\AcaOrBc{}$ after removing the now unused b-transition.}\label{fig:example-simplify-2}
  \end{minipage}
  \end{figure}

  Plugging in the same inequality again on the same principle, we get $2(1 -
  l_c') > l_c' \equiv 1 > l_c'$ since we are working with integers, and
  therefore that $l_c' = 0$. This gives us the even smaller automaton of
  Figure~\ref{fig:example-simplify-3}.

  \begin{figure}[h]
    \begin{minipage}[b]{0.75\linewidth}
    \centering 
      \includegraphics[width=\textwidth]{aca_or_bc_simplified_3}
      \caption{$\AcaOrBc{}$ after further constraint propagation.}\label{fig:example-simplify-3}
    \end{minipage}
    \end{figure}

This will produce a straightforward product since both automata now have the
same shape, where we essentially only need to expand the self-loops of
$\SomethingCSomething$ and reduce its range labels to only capture a:s. Note
that we got this result by only performing linear inequality reasoning on the
automata of the product versus the number of times each transition would be
used.
